---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r packages}
require(knitr)
require(ggplot2)
```

```{r loaddata}
if(!file.exists("activity.csv")) {
        unzip("activity.zip")
}
activity <- read.csv("activity.csv")
```

```{r formatdate}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
```{r aggregatesteps}
aggSteps <- tapply(activity$steps,activity$date,sum,na.rm=TRUE)
```

```{r stepshistogram}
qplot(aggSteps, bins = 9, xlab = "Number of Steps", main = "Total Number of Steps Taken in a Day")
```

```{r avesteps}
meanSteps <- mean(aggSteps)
medianSteps <-median(aggSteps)
```

The mean number of steps taken per day is `r meanSteps` and the median number of
steps taken per day is `r medianSteps`.

## What is the average daily activity pattern?
```{r intervalsteps}
intSteps <- tapply(activity$steps,activity$interval,mean,na.rm=TRUE)
```

```{r plotsteps}
plot(names(intSteps),intSteps,type = "l", col = "blue", 
     xlab = "Interval (minutes)", ylab = "Number of Steps", 
     main = "Average Number of Steps in Each 5 Minute Inerval")
```

```{r maxinterval}
maxSteps <- max(intSteps)
maxInt <- names(intSteps[which(intSteps == maxSteps)])
```

The interval with the most average steps is `r maxInt` during which an average
of `r maxSteps` are taken.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
