---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r options}
options(scipen = 1, digits = 2)
```

This first line simply insures that any numbers that may be printed using
scientific notation are printed in standard form instead.

```{r packages}
require(knitr)
require(ggplot2)
require(dplyr)
```

Here we have all packages that are required for the analysis.

```{r loaddata}
if(!file.exists("activity.csv")) {
        unzip("activity.zip")
}
activity <- read.csv("activity.csv")
```

Now we begin to actually load the data. We check to see if the file has already
been unzipped, and if it has not, we unzip the file. Then the file is read into
the `activity` variable using `read.csv()`.

```{r formatdate}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

Lastly, since it will be important in order to answer future questions, we
reformat the `date` column of the `activity` data frame to be in the date
format.

## What is mean total number of steps taken per day?
```{r aggregatesteps}
aggSteps <- tapply(activity$steps,activity$date,sum,na.rm=TRUE)
```

First, we can use `tapply()` to sum up the number of steps in each day, removing
any `NA`'s for now.

```{r stepshistogram}
qplot(aggSteps, bins = 9, xlab = "Number of Steps", 
      main = "Total Number of Steps Taken in a Day", col = I("black"),
      fill = I("red"))
```

When given only one variable, `qplot()` creates a histogram. 9 bins were chosen
in order to give a clear overall picture of the data. Here we see quite a few
days in the lowest bin, likely due to days with many or all `NA`'s.

```{r avesteps}
meanSteps <- mean(aggSteps)
medianSteps <-median(aggSteps)
```

The mean number of steps taken per day is `r meanSteps` and the median number of
steps taken per day is `r medianSteps`.

## What is the average daily activity pattern?
```{r intervalsteps}
intSteps <- tapply(activity$steps,activity$interval,mean,na.rm=TRUE)
```

Here, `tapply()` is used again, but this time to average the number of steps
and this time based on the interval instead of the day (again removing `NA`'s).

```{r plotsteps}
qplot(as.numeric(names(intSteps)),intSteps, 
     xlab = "Interval (minutes)", ylab = "Number of Steps", 
     main = "Average Number of Steps in Each 5 Minute Inerval") + 
        geom_line(col = "blue", size = 2)
```

`tapply()` gives an array where we have a vector of averages, and each element
is given its respective interval as a name. Thus the names of `intSteps` are
used as the independent variable and the calculated means are used as the
dependent variable to create this time series graph.

```{r maxinterval}
maxSteps <- max(intSteps)
maxInt <- names(intSteps[which(intSteps == maxSteps)])
```

The interval with the most average steps is `r maxInt` during which an average
of `r maxSteps` steps are taken.

## Imputing missing values
```{r nacount}
colSums(is.na(activity))
countNA <- sum(is.na(activity$steps))
```

There are `r countNA` observations that have an `NA` value listed. We see that
the `steps` column is the only one that has any `NA` values at all.

```{r replacena}
modAct <- activity$steps
modAct[c(1,17568)] <- intSteps[c(1,288)]
for (i in 2:17567){
        if (is.na(modAct[i])){
                near <- mean(c(activity$steps[i-1], activity$steps[i+1]))
                if (is.na(near))
                        modAct[i] <- intSteps[which(activity$interval[i] == names(intSteps))]
                else
                        modAct[i] <- mean(c(near, intSteps[which(activity$interval[i] == names(intSteps))]))
        }
}
newAct <- activity
newAct$steps <- modAct
```

In order to account for activity in intervals nearby to a missing value and for
usual movement at a given time of day, we find the mean of the intervals
immediately before and after the missing interval and find the mean of the
given interval across all days in the dataset. We then take the mean of these
two numbers to replace the `NA` value with. If both values around the interval
are also `NA`, the value is computed from the daily interval mean only.  
NOTE: Since the code inside the `for` loop will be invalid for `i in c(1,288)`, 
those two spots are manually calculated outside of the `for` loop. Each of the 
values are surrounded by `NA`'s and thus are recalculated using the mean of 
their respective intervals.

```{r aggregateallsteps}
newAggSteps <- tapply(newAct$steps,newAct$date,sum,na.rm=TRUE)
```

```{r allstepshistogram}
qplot(newAggSteps, bins = 9, xlab = "Number of Steps", 
      main = "Total Number of Steps Taken in a Day (NA's Imputed", 
      col = I("black"), fill = I("red"))
```

These two steps are equivalent to those taken before, just with all of the `NA`
values having been imputed. Here we see fewer values in the lowest bin since the
days that were mostly or all `NA` have now been replaced by actual values. This
gives the distribution a more normal shape.

```{r aveallsteps}
meanNewSteps <- mean(newAggSteps)
medianNewSteps <-median(newAggSteps)
```

The mean number of steps taken per day is `r meanNewSteps` and the median number
of steps taken per day is `r medianNewSteps`. It is clear that both the mean and
the median number of steps has increased when replacing the `NA` values. 
Obviously adding in more values will only increase the estimates of total daily 
number of steps.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays}
newAct <- mutate(newAct,
                 type = as.factor(ifelse(weekdays(date) %in% c("Saturday","Sunday"), "weekend", "weekday")))
```

First, a new factor variable is created that checks which day of the week the
observation ocurred during and assigns the new variable a level of either
`"weekend"` or `"weekday"`.

```{r weekdayaverages}
weekdayAct <- filter(newAct, type == "weekday")
weekendAct <- filter(newAct, type == "weekend")
weekdayInt <- tapply(weekdayAct$steps, weekdayAct$interval, mean)
weekendInt <- tapply(weekendAct$steps, weekendAct$interval, mean)
```

First, the `newAct` data frame is filtered into two new data frames based on
type of day (weekday/weekend). Then, just as before, the number of steps are
averaged by interval.

```{r weekdayplots}
par(mfrow = c(2,1), mar = c(2,1,1,1), oma = c(4,4,2,0))
rng <- range(c(weekendInt,weekdayInt))
plot(as.numeric(names(weekendInt)), weekendInt, xlab = "", ylab = "",
     main = "Weekend", type = "l", col = "blue", ylim = rng)
plot(as.numeric(names(weekdayInt)), weekdayInt, xlab = "", ylab = "",
     main = "Weekday", type = "l", col = "blue", ylim = rng)
mtext("Interval", side = 1, outer = TRUE, line = 0)
mtext("Number of steps",side = 2, outer = TRUE, line = 2)
```

The data is then plotted as a time series as before, using the `rng` variable
to insure that each plot has the same `ylim` argument. This allows for an easier
comparison between the two plots.